---
title: "Visualizing Police-Public Contact and Crime Victimization"
author: "Mandi Acevedo and Bela Walkin"
format: html
self-contained: true
editor: visual
---

```{r}
library(haven)
PPCS <- read_dta("PPCS.dta")
library(tidyverse)
```

# Visualization 1

```{r}

PPCS |>
  mutate(as.numeric(ARRESTED)) |>
  filter(ARRESTED == 1) |>
  ggplot(aes(x = EDUCATION_SUB)) +
  geom_histogram(binwidth=1,
                 fill = "pink",
                 color = "black") +
  scale_x_continuous(breaks = seq(0, 18, 1)) +
  theme_minimal() + 
  labs(title = paste0("Of survey respondents who have been arrested, most have 
no more than a high school level education."),
      subtitle = "Includes responses from 2002, 2008, 2011",
      caption = "Source: Police-Public Contact Survey and National Crime Victimization Survey, Accessed Through ICPSR", 
      x = "Years of Education",
      y = "Number of Respondents")
```
## Understanding who comes into contact with police, and subsequently gets arrested, is extremely improtant when researching how to lower police-citizens contact that results in arrests. Of those individuals in the Police-Citizen contact survey, most individuals have no more than a high school diploma (12 yeas of education). This demonstrates that those interacting with the jsutice system in negative ways may be more likely to stop education earlier. 

# Visualization 2

```{r}

PPCS |>
  filter(NUM_CONT < 75) |>
  ggplot() +
  geom_point(aes(x = NUM_MOVES,
                 y = NUM_CONT),
             alpha = 0.1,
             color = "pink") +
  geom_smooth(mapping = aes(x = NUM_MOVES,
                            y = NUM_CONT),
              method = "lm",
              se = FALSE,
              color = "black") +
  scale_x_continuous(breaks = seq(0, 75, 5)) +
  theme_minimal() +
  labs(title = paste("Frequent moving is asociated with more police contacts."),
       subtitle = "Survey Respondents from 2002, 2008, 2011",
       caption = paste("Source: Police-Public Contact Survey and", 
"National Crime Victimization Survey, Accessed Through ICPSR"),
       x = "Number of times moved in the last 5 years",
       y = "Amount of police contacts in past 12 months")
```
## Of individuals who have had less than 75 contatcs with police officers in the last 12 months (non-outliers), frequent moving is associated with higher police contacts. Individuals who have less stable home-bases are more likley to have contacts with police. This could mean that moving leads to higher liklihood of victimization or increased criminal-behavior for an individual, as some examples.

# Visualization 3

```{r}
PPCS |>
  mutate(TOTAL_VALUE_LOSS = VAL_LOSS2_INC_P23HH + VAL_LOSS2_INC_P23PER) |>
  mutate(MSA_STATUS = case_when(
    MSA_STATUS == 1 ~ "Urban",
    MSA_STATUS == 2 ~ "Suburban",
    MSA_STATUS == 3 ~ "Rural")) |>
  ggplot(aes(
    x = factor(MSA_STATUS), 
    y = TOTAL_VALUE_LOSS))+
  geom_boxplot(outlier.shape = NA,
               fill = "pink",
               color = "black")+
  scale_y_continuous(limits = c(0,750)) +
  theme_minimal()+
  labs(title = "Stolen property values are, on average, highest amongst suburban victims.",
         subtitle = paste("Distribution of reported value of cash/property lost",
         "by Metropolitan Statistical Area (MSA) type"),
         caption = "Source: Police-Public Contact Survey and National Crime Victimization Survey, Accessed through ICPSR",
         x = "MSA type",
         y = "Total value of cash & property lost (USD)")

```

# Visualization 4

```{r}
PPCS |>
  mutate(CONTACT = ifelse(NUM_CONT > 0, 1, 0),
         C4_RACE = case_when(
           C4_RACE == 1 ~ "White",
           C4_RACE == 2 ~ "Black",
           C4_RACE == 3 ~ "Hispanic",
           C4_RACE == 4 ~ "Other / Multiracial")) |>
  group_by(C4_RACE) |>
  filter(PPCS_YEAR == 2011) |>
  summarize(
    total = n(),
    contact = replace_na(
      sum(CONTACT == 1),0),
    no_contact = replace_na(
      (total - contact),0)) |>
  pivot_longer(
    cols = c(contact, no_contact),
    names_to = "category",
    values_to = "count") |>
  mutate(category = ifelse(category == "contact","yes","no")) |>
  ggplot(aes( 
    x = reorder(C4_RACE, total),
    y = count,
    fill = category)) +
  geom_bar(stat="identity") +
  geom_text(aes(label = count),
            position = position_stack(vjust = 0.5),
            size = 3) +
  scale_fill_manual(values = c("yes" = "pink", "no" = "lightgreen")) +
  labs(title = paste("White individuals are less likely to have had",
                     "police contact"),
       subtitle = paste("Proportion of 2011 respondents who have had at least one",
                        "face-to-face contact with police in past 12 months"),
       caption = paste("Source: Police-Public Contact Survey and National",
                       "Crime Victimization Survey, Accessed through ICPSR"),
       y = "Respondent count",
       fill = "Ever had police contact?") +
  theme_minimal() +
  theme(axis.title.x = element_blank())
```
